Case-Base Neural Network: survival analysis with time-varying, higher-order interactions

### 2023-05-18 StatPaper/CBNNManuscript/CBNNManuscript.tex
StatSubmission

\documentclass[APA,LATO1COL]{WileyNJD-v2}
\usepackage{lmodern}

@@

\corres{Jesse Islam's \email{jesse.islam@mail.mcgill.ca}}

@@

\abstract[Summary]{Neural network-based survival methods can model data-driven covariate interactions. While these methods have led to better predictive performance than regression-based approaches, they cannot model time-varying interactions and complex baseline hazards simultaneously. To address this, we propose Case-Base Neural Network (CBNN) as a new approach that combines the case-base sampling framework with flexible neural network architectures. Using a novel sampling scheme and data augmentation to naturally account for censoring, we construct a feed-forward neural network that may take time as an input. CBNN predicts the probability of an event occurring at a given moment to estimate the hazard function. We compare the performance of CBNN to neural network and regression-based survival methods in a simulation and three case studies using two time-dependent metrics. First, we examine performance on a simulation involving a complex baseline hazard and time-varying interactions to assess all methods, with CBNN outperforming competitors. Then, we apply all methods to two real data applications where CBNN outperforms the competing models as well. In a third case study, CBNN shows competitive performance. Our results highlight the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of single event survival outcomes. An R package is available at https://github.com/Jesse-Islam/cbnn.}

@@

A common assumption we use in survival analysis is that the risk of the event of interest does not vary with
time (i.e. proportional hazards) \citep{hanley2009}. This simplifying assumption results in using Cox proportional
hazards models more often than Smooth-in-time, accelerated failure time (AFT) models \citep{hanley2009}. This
preference causes analyses to be based on hazard ratios and relative risks rather than on survival curves and
absolute risks \citep{hanley2009}. The proportional hazards assumption in a Cox model may be incorrect in studies
where the disease pathogenesis may change over time \citep{coradini2000time}. For example, studies of breast cancer
incidence have shown that the risk of developing a tumor involves a time-varying interaction with tumor size
\citep{coradini2000time}. Time-varying covariates can be incorporated easily into AFT models \citep{royston2002flexible}.
To provide flexibility in the baseline hazard, we can use a basis of splines on time as model inputs
\citep{royston2002flexible}. However, these regression-based models require prior knowledge of potential
time-varying interactions and their quantitative effects.

Neural networks provide a data-driven approach to approximating interaction terms in models. For example.
DeepSurv is a neural network-based proportional hazards model that implements the Cox partial log-likelihood
as a custom loss function \citep{katzman2018DeepSurv}. This results in a stepwise absolute risk curve that cannot
accommodate time-varying interactions. Compared to Cox regression, DeepSurv shows better performance on a
real dataset \citep{katzman2018DeepSurv}. Previous work suggests we may change the loss function to address
non-proportional hazards \citep{faraggi1995neural}. In contrast, DeepHit assumes an inverse Gaussian distribution
as the baseline hazard \citep{lee2018DeepHit} and specifies each follow-up time of interest in the model. DeepHit
directly estimates survival curves, rather than deriving a hazard function \citep{lee2018DeepHit}. On real datasets,
DeepHit outperformed DeepSurv \citep{lee2018DeepHit}. We note that DeepSurv and DeepHit both use custom loss
functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}. In addition, DeepHit introduces a hyperparameter
weighing its two loss functions (negative log-likelihood and ranking losses) \citep{lee2018DeepHit}. Compared
to these neural network methods, regression-based approaches require prior specification of all interaction terms,
which makes it challenging to model covariate effects that change with time. The current neural network models
provide flexibility at the cost of clarity, while regression models provide clarity at the cost of flexibility.

We propose Case-Base Neural Networks (CBNN) as a method for single event survival analysis that models
time-varying interactions and a flexible baseline hazard. CBNN use the case-base sampling technique, which
allows probabilistic models to predict survival outcomes \citep{hanley2009}. After case-base sampling, we
implement a model using common neural network components that uses transformations of time as a feature
(covariate) to specify different baseline hazards. For example, by including splines of time as covariates,
case-base with logistic regression (CBLR) can approximate the Royston-Parmar flexible baseline hazard
\citep{royston2002flexible} \citep{hanley2009}. However, this still requires explicit specification of
time-varying interactions. CBNN can model both without extra tuning parameters.

In this paper, we describe the case-base sampling method and compare its properties to neural
network models in Section \ref{methods}, along with our hyperparameter selection procedure,
metrics and software implementation. We also explore the performance of CBNN, DeepSurv,
DeepHit, Cox regression and CBLR on simulated data in Section \ref{sims}. Section \ref{casestudies}
describes the performance of the same methods in three case studies. Section \ref{discussion} explores
the implications of our results and contextualizes them within neural network survival analysis in a single
event setting.

@@

\hypertarget{methods}{%
\section{Case-base neural networks, metrics, hyperparameters and
software}\label{methods}}

Case-base sampling divides total survival time for each individual into person-specific moments
(person-moments), treating each person-moment as a sample. In this section, we detail how neural
networks explicitly incorporating time as a feature while adjusting for the sampling bias in case-base
sampled data; and describe the metrics and hyperparameter selection procedure used to compare
CBNN with other software packages. An R package to use CBNN is available at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to reproduce the figures
and empirical results in this paper is available at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

Case-base sampling is an alternative framework for survival analysis \citep{hanley2009},
which converts the total survival time into discrete person-moments. Without this preliminary step,
we cannot model survival outcomes with probabilistic methods. In case-base sampling, we sample
from the continuous survival time of each person in our dataset to create a \emph{base series} of
\emph{person-moments}. This \emph{base series} complements the \emph{case series}, which contains
all person-moments at which the event of interest occurs.

For each person-moment sampled, let \(X_i\) be the corresponding covariate profile
\(\left(x_{i1},x_{i2},...,x_{ip} \right)\), \(T_i\) be the time of the person-moment and \(Y_i\) be the indicator
variable for whether the event of interest occurred at time \(T_i\). We estimate the hazard function
\(h(t \mid X_i)\) using the sampled person-moments. Recall that \(h(t \mid X_i)\) is the instantaneous
potential of experiencing the event at time \(t\) for a given set of covariates \(X_i\), assuming \(T_i \geq t\).

Now, let \(b\) be the (user-defined) size of the \emph{base series} and let \(B\) be the sum of all follow-up
times for the individuals in the study. Let $c$ be the number of events in the \emph{case series}. A
reasonable concern is determining how large $b$ should be relative to $c$, since the size of \(b\)
influences how much information is lost in the sampling process \citep{hanley2009}. Using our notation,
Hanley $\&$ Miettinen (2009) adapt a quote from Mantel (1973): "By the reasoning that
$\frac{cb}{c+b}=\left[\left( \frac{1}{c}+\frac{1}{b}\right)]\right]$ measures the relative information in a comparison of two
averages based on sample sizes of c and b respectively, we might expect by analogy, which would
of course not be exact in the present case, that this approach would result in only a moderate loss of
information. (The practicing statistician is generally aware of this kind of thing. There is little to be gained
by letting the size of one series, b, become arbitrarily large if the size of the other series, c, must remain
fixed)" \citep{hanley2009} \citep{mantel1}. If $b=100c$, then we expect learned weights to be at most one
percent higher than if the entire study base $B$ was used, as they are proportional to
$\frac{1}{c}+\frac{1}{100c}$ rather than $\frac{1}{c}+\frac{1}{\infty c}$ \citep{hanley2009}.

If we sample the \emph{base series} uniformly across the study base, then the hazard function
of the sampling process is equal to \(b/B\). Therefore, we have the following equality \cite{saarela2015}
\footnote{For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

@@

After case-base sampling, we pass all features, including time, into any user-defined feed-forward
component. Then, we add an offset term and pass the output through a sigmoid activation function
(Figure \ref{fig:NNarch}). Since we are interested in predicting the odds of an event occurring, the sigmoid
activation function is ideal as it is the inverse of the odds and can be used to calculate the hazard. The
general form for the neural network using CBNN is:

@@

where \(T\) is a random variable representing the event time, \(X\) is the random variable for a covariate
profile, \(f_{\theta}(X, T)\) represents any feed-forward neural network architecture, \(\log\left(\frac{B}{b}\right)\)
is the offset term for adjust for the bias (\(\log\left(\frac{b}{B}\right)\)) set by case-base sampling, \(\theta\) is the
set of parameters learned by the neural network and \(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating
a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven,
where user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility
of the hazard function.

@@

\caption{Methodological steps involved in CBNN. The first step, case-base sampling, is completed before training begins.
We pass this sampled data through a feed-forward neural network and add an offset to adjust for the bias inherent in
case-base sampling, then apply a sigmoid activation function to create a probability. Once the neural network model
completes its training, we can convert the probability output to a hazard for the survival outcomes of interest.}\label{fig:NNarch}
\end{figure}

@@

Backpropagation with an appropriate minimization algorithm (e.g.~Adam, RMSPropagation, stochastic gradient
descent) is used to optimize the parameters in the model \citep{gulli2017}. For our analysis, we use Adam as implemented
in Keras \citep{gulli2017}. While the size of the \emph{case series} is fixed as the number of events; the size of the
\emph{base series} is not restricted. We use a ratio of 100:1 \emph{base series} to \emph{case series} \citep{hanley2009}.
We pass our feed-forward neural network through a sigmoid activation function (Figure \ref{fig:NNarch}). Finally, we can
convert this model output to a hazard. To use CBNN for predictions, we manually set the offset term \(\log\left(\frac{B}{b} \right)\)
to 0 in the new data as we already account for the sampling bias during the fitting process.

Since we are directly modeling the hazard, we can readily estimate the
risk function (\(F\)) at time \(t\) for a covariate profile \(X\),
viz.\begin{align}\label{eqn:ci2}
F\left(t\mid X\right)& = 1 - \exp\left(-\int_{0}^{t}h(u|X) \,\textrm du\right).
\end{align} We use a finite Riemann sum \citep{hughes2020calculus} to
approximate the integral in \eqref{eqn:ci2}.

@@

To choose our method-specific hyperparameters, we used the Integrated Brier Score (IBS) \citep{graf1999}, which is based on the
Brier Score (BS) and provides a summarized assessment of performance for each model. We assess the performance of models on
a held out dataset using two metrics: 1) the Index of Prediction Accuracy (IPA) \citep{kattan2018index}; and 2) the Inverse probability
censoring weights-adjusted time dependent area under the receiver operating characteristic curve ($AUC_{IPCW}$) \citep{auc}. The
IPA score is a metric for both discrimination and calibration, while the $AUC_{IPCW}$ provides a metric for discrimination only.

\hypertarget{bs}{%
\subsubsection{Brier Score (BS)}\label{bs}}
The BS \citep{graf1999} is defined as \begin{align}\label{eqn:bs}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
\end{align} where \(\delta_{i}=1\) shows individuals who have experienced the event, \(N\) represents the number of samples
in our dataset over which we calculate \(BS(t)\), \(\widehat{G}(t)=P[c>t]\) is a non-parametric estimate of the censoring distribution,
\(c\) is censoring time and \(T_{i}\) is an individual's survival or censoring time. The BS provides a score that accounts for the
information loss because of censoring. Once we fix our \(t\) of interest, the individuals in the dataset can be divided into three
groups. Individuals who experienced the event before \(t\) are present in the first term of the equation. The second term of the
equation includes individuals who experience the event or are censored after \(t\). Those censored before \(t\) (the remaining
individuals) are accounted for by the IPCW adjustment (\(G(\cdot)\)) present on both terms.

@@

The Integrated Brier Score (IBS) is a function of the BS \ref{eqn:bs} \citep{graf1999}, which is defined as \begin{align*}
IBS(t)=\int_{t}^{t_{max}}BS(t)dw(t),
\end{align*} where $w(t)=\frac{t}{t_{max}}$. As the IBS is calculated for a range of follow-up-times,
it is a useful metric to track overall performance across all of follow-up time during cross-validation. We use the
IPA score to get an estimate of performance at each follow-up time for our studies.

@@

The IPA as a function of time, based on the BS. The IPA score is given by \begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where \(BS_{model}(t)\) represents the BS over time \(t\) for the model of interest and \(BS_{null}(t)\) represents
the BS if we use an unadjusted Kaplan-Meier (KM) curve as the prediction for all observations \citep{kattan2018index}. The
IPA score has an upper bound of one, where positive values show an increase in performance over the null model and
negative values show that the null model performs better. An extension of BS, the IPA scores assess both model calibration
and discrimination and show how performance changes over follow-up time \citep{graf1999} \citep{kattan2018index}.

@@

R \citep{Rsoft} and Python \citep{py} are used to evaluate methods from both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg} and the CBLR model using the \textbf{casebase} package \citep{cbpkg}. Both DeepSurv and DeepHit are fit using \textbf{pyCox}
\citep{lee2018DeepHit}. We made the components of CBNN using the \textbf{casebase} package \citep{cbpkg} for the sampling step and the
\textbf{keras} \citep{keras} package for our neural network architecture. We use the \textbf{simsurv} package \citep{simsurv} for our simulation studies
and \textbf{flexsurv} \citep{flexsurv} to fit a flexible baseline hazard using splines for our complex simulation. The \textbf{riskRegression} package
\citep{riskRegression} is used to get the IPA and $AUC_{IPCW}$. We modify the \textbf{riskRegression} package to be used with any user supplied
risk function \(F\). We use the \textbf{reticulate} package \citep{reticulate} to run both R based methods and python based methods on the same seed.

@@

\caption{Performance of each model in the complex simulation (A, E), multiple myeloma (MM) case study (B, F), free light chain (FLC) case study (C, G)
and prostate cancer (Prostate) case study (D, H). The first row shows the IPA for each model in each study over follow-up time. Negative values mean
the model performs worse than the null model and positive values mean the model performs better. The second row shows the $AUC_{IPCW}$ for
each model in each study over follow-up time, where higher is better. Each model-specific metric in each study shows a 95\% confidence interval over 100 iterations. Metrics
are shown for six models: Case-Base with Logistic Regression (CBLR), Case-Base Neural Networks (CBNN), Cox Proportional Hazard (Cox), DeepHit
and DeepSurv. The Kaplan-Meier (KM) model serves as a baseline, predicting the average curve for all individuals. CBLR and Cox have near identical
performance, resulting in curves that overlap. The Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) model shows
the best performance we can expect on the simulated data.}\label{fig:megaPlot}
\end{figure}

@@

We simulate data to evaluate the performance of CBNN in comparison with existing regression (Cox, CBLR) and neural network (DeepHit, DeepSurv)
methods. We specify a linear combination of each covariate as the linear predictor in the regression-based methods (Cox, CBLR), which contrasts
with neural network approaches that allow for non-linear interactions. The simulated data are based on a complex baseline hazard with time-varying
interactions, each with 10\% random censoring. We simulate three covariates for 5000 individuals:

\[
z_{1} \sim \textrm{Bernoulli}(0.5) \qquad \qquad %technically a control and treatment group
z_{2} \sim \begin{cases}
 N(0,0.5) & \textrm{if } z_{1}=0\\ 
 N(1,0.5) & \textrm{if } z_{1}=1
\end{cases} \qquad \qquad
z_{3} \sim N(1,0.5).
\]
In addition to the methods described above, we include the exact functional form of the covariates in a CBLR model (referred to as Optimal for simplicity)
in the complex simulation. We conduct 100 bootstrap re-samples on the training data to get confidence intervals. We kept 15\% of the data for testing before
hyperparameter selection. 15\% of the remaining data is used for validation and the rest is reserved for training. We predict risk functions for individuals in
the test set, which are used to calculate our IPA and $AUC_{IPCW}$.

@@

Figure \ref{fig:megaPlot} A, E and Table \ref{tab:megaTable} shows the performance over time on a test set. For discrimination, the Optimal model performs best,
followed by CBNN, DeepHit, DeepSurv and the linear models (Figure \ref{fig:megaPlot} E). For both discrimination and calibration, the rankings remain the same
except for DeepHit which shows a periodic loss in performance (Figure \ref{fig:megaPlot} A). The Optimal model acts as a reference for ideal performance on the
simulated data, with CBNN performing better than the competitors in the complex simulation. To obtain a more realistic performance assessment, we compared
models in three case studies with a time-to-event outcome.

>>>>

\documentclass[APA,LATO1COL]{CBNNManuscript}
\usepackage{lmodern}

@@

\corres{Jesse Islam \email{jesse.islam@mail.mcgill.ca}}

@@

\abstract[Summary]{Neural network-based survival methods can model data-driven covariate interactions. While these methods can provide better predictive performance than regression-based approaches, not all can model time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNNs) as a new approach that combines the case-base sampling framework with flexible neural network architectures. Using a novel sampling scheme and data augmentation to naturally account for censoring, we construct a feed-forward neural network that may take time as an input. CBNNs predict the probability of an event occurring at a given moment to estimate the hazard function. We compare the performance of CBNNs to regression and neural network-based survival methods in a simulation and three case studies using two time-dependent metrics. First, we examine performance on a simulation involving a complex baseline hazard and time-varying interactions to assess all methods, with CBNN outperforming competitors. Then, we apply all methods to three real data applications, with CBNNs outperforming the competing models in two studies and showing similar performance in the third. Our results highlight the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of single event survival outcomes. An R package is available at https://github.com/Jesse-Islam/cbnn.}

@@

A common assumption in survival analysis is that the risk of the event of interest does not vary with
time (i.e. proportional hazards) \citep{hanley2009}. This simplifying assumption results in using Cox proportional
hazards models more often than smooth-in-time, accelerated failure time (AFT) models \citep{hanley2009}. This
preference causes analyses to be based on hazard ratios and relative risks rather than on survival curves and
absolute risks \citep{hanley2009}. The proportional hazards assumption in a Cox model may be incorrect in studies
where the disease pathogenesis may change over time \citep{coradini2000time}. For example, studies of breast cancer
incidence have shown that the risk of developing a tumor involves a time-varying interaction with tumor size
\citep{coradini2000time}. These time-varying covariates can be incorporated easily into AFT models \citep{royston2002flexible}.
However, these regression-based models require prior knowledge of potential
time-varying interactions and their quantitative effects.

Neural networks provide a data-driven approach to approximating interaction terms that benefit prediction. For example,
DeepSurv is a neural network-based proportional hazards model that implements the Cox partial log-likelihood
as a custom loss function \citep{katzman2018DeepSurv}. This results in a stepwise absolute risk curve that cannot
accommodate time-varying interactions. DeepSurv has shown better performance than Cox proportional hazard models in the Rotterdam and German Breast Cancer Study Group and The Molecular Taxonomy of Breast Cancer International Consortium datasets \citep{katzman2018DeepSurv}. It has also been used to develop personalized prediction models for cancers in a number of studies \citep{ds1} \citep{ds2} \citep{ds3}. Previous work suggests we may change the loss function to address
non-proportional hazards \citep{faraggi1995neural}. In contrast, DeepHit assumes an inverse Gaussian distribution
as the baseline hazard \citep{lee2018DeepHit} and specifies each follow-up time of interest in the model. DeepHit
directly estimates survival curves, rather than deriving a hazard function \citep{lee2018DeepHit}. On real datasets,
DeepHit outperformed DeepSurv \citep{lee2018DeepHit}. We note that DeepSurv and DeepHit both use custom loss
functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}. In addition, DeepHit introduces a hyperparameter
weighing its two loss functions (negative log-likelihood and ranking losses) \citep{lee2018DeepHit}. Compared
to these neural network methods, regression-based approaches require prior specification of all interaction terms,
which makes it challenging to model covariate effects that change with time. The current neural network models
provide flexibility at the cost of clarity, while regression models provide clarity at the cost of flexibility.

We propose Case-Base Neural Networks (CBNNs) as a method for single event survival analysis that models
time-varying interactions and a flexible baseline hazard. CBNNs use the case-base sampling technique, which
allows probabilistic models to predict survival outcomes \citep{hanley2009}. After case-base sampling, we
implement a model using common neural network components that uses transformations of time as a feature
(covariate) to specify different baseline hazards. For example, by including splines of time as covariates,
case-base with logistic regression (CBLR) can approximate the Royston-Parmar flexible baseline hazard
\citep{royston2002flexible} \citep{hanley2009}. However, this still requires explicit specification of
time-varying interactions. CBNN can model both without extra tuning parameters.

In this paper, we describe the case-base sampling method and compare its properties to neural
network models, along with our hyperparameter selection procedure,
metrics and software implementation. Next, we explore the performance of CBNN, DeepSurv,
DeepHit, Cox regression and CBLR on simulated data and
describe their performance in three case studies. Finally, we explore
the implications of our results and contextualize them within neural network survival analysis in a single
event setting.

@@

\hypertarget{methods}{%
\section{Case-base neural network, metrics, hyperparameters and
software}\label{methods}}

Case-base sampling is an alternative framework for survival analysis \citep{hanley2009},
which converts the total survival time into discrete person-moments. In this section, we detail how neural
networks explicitly incorporate time as a feature while adjusting for the sampling bias in case-base
sampled data and describe the metrics and hyperparameter selection procedure used to compare
CBNN with other software packages. An R package to use CBNN is available at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to reproduce the figures
and empirical results in this paper is available at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

To implement case-base sampling, we divide the total survival time for each individual into discrete person-specific moments (person-moments) and treat each person-moment as a sample. This creates a \emph{base series} of
\emph{person-moments} where an event does not occur. This \emph{base series} complements the \emph{case series}, which contains
all person-moments at which the event of interest occurs.

For each person-moment sampled, let \(X_i\) be the corresponding covariate profile
\(\left(x_{i1},x_{i2},...,x_{ip} \right)\), \(T_i\) be the time of the person-moment and \(Y_i\) be the indicator
variable for whether the event of interest occurred at time \(T_i\). We estimate the hazard function
\(h(t \mid X_i)\) using the sampled person-moments. Recall that \(h(t \mid X_i)\) is the instantaneous
risk of experiencing the event at time \(t\) for a given set of covariates \(X_i\), assuming \(T_i \geq t\).

Now, let \(b\) be the (user-defined) size of the \emph{base series} and let \(B\) be the sum of all follow-up
times for the individuals in the study. Let $c$ be the number of events in the \emph{case series}. A
reasonable concern is determining how large $b$ should be relative to $c$, since the size of \(b\)
influences how much information is lost in the sampling process \citep{hanley2009}. The relative information when comparing two averages
is measured by $\frac{cb}{c+b}=\left[\left( \frac{1}{c}+\frac{1}{b}\right)\right]$, where $b$ is the sample size of the \emph{base series} and $c$
is the sample size of the \emph{case series}  \citep{hanley2009} \citep{mantel1}. If $b=100c$, then we expect learned weights to be at most one
percent higher than if the entire study base $B$ was used, as they are proportional to
$\frac{1}{c}+\frac{1}{100c}$ rather than $\frac{1}{c}+\frac{1}{\infty c}$ \citep{hanley2009} \citep{mantel1}.

If we sample the \emph{base series} uniformly across the study base, then the hazard function
of the sampling process is equal to \(b/B\). Therefore, we have the following equality \citep{saarela2015}
[For a rigorous treatment, see Saarela \& Hanley (2015) section 3].:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being in the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

@@

After case-base sampling, we pass all features, including time, into any user-defined feed-forward
component (Figure \ref{fig:NNarch}). Then, we add an offset term and pass the output through a sigmoid activation function
(Figure \ref{fig:NNarch}). Since we are interested in predicting the odds of an event occurring, the sigmoid
activation function is ideal as it is the inverse of the odds and can be used to calculate the hazard. The
general form for the neural network using CBNN is:

@@

where \(T\) is a random variable representing the event time, \(X\) is the random variable for a covariate
profile, \(f_{\theta}(X, T)\) represents any feed-forward neural network architecture, \(\log\left(\frac{B}{b}\right)\)
is the offset term to adjust for the bias (\(\log\left(\frac{b}{B}\right)\)) set by case-base sampling, \(\theta\) is the
set of parameters learned by the neural network and \(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating
a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven,
while user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility
of the hazard function.

@@

\caption{Methodological steps involved in CBNN. The first step, case-base sampling, is completed before training begins.
Then, we pass this sampled data through a feed-forward neural network and add an offset to adjust for the bias inherent in
case-base sampling and apply a sigmoid activation function to create a probability. Once the neural network model
completes its training, we can convert the probability output to a hazard for the survival outcome of interest.}\label{fig:NNarch}
\end{figure}

@@

Backpropagation with an appropriate minimization algorithm (such as Adam, RMSPropagation, stochastic gradient
descent) is used to optimize the parameters in the model \citep{gulli2017}. For our analysis, we use Adam as implemented
in Keras \citep{gulli2017}. While the size of the \emph{case series} is fixed as the number of events; the size of the
\emph{base series} is not restricted. We use a ratio of 100:1 \emph{base series} to \emph{case series} \citep{hanley2009}.
After fitting our model, we
convert the output to a hazard. To use CBNN for predictions, we manually set the offset term $\left(\log\left(\frac{B}{b} \right)\right)$
to 0 in the new data as we already account for the sampling bias during the fitting process.

Since we are directly modeling the hazard, we can readily estimate the
risk function (\(F\)) at time \(t\) for a covariate profile \(X\),
\begin{align}\label{eqn:ci2}
F\left(t\mid X\right)& = 1 - \exp\left(-\int_{0}^{t}h(u|X) \,\textrm du\right).
\end{align} We use a finite Riemann sum \citep{hughes2020calculus} to
approximate the integral in \eqref{eqn:ci2}.

@@

To choose our method-specific hyperparameters, we use the Integrated Brier Score (IBS) \citep{graf1999}, which is based on the
Brier Score (BS) and provides a summarized assessment of performance for each model. We assess the performance of models on
a held out dataset using two metrics: 1) the Index of Prediction Accuracy (IPA) \citep{kattan2018index}; and 2) the Inverse probability
censoring weights-adjusted time dependent area under the receiver operating characteristic curve ($AUC_{IPCW}$) \citep{auc}. The
IPA score is a metric for both discrimination and calibration, while the $AUC_{IPCW}$ provides a metric for discrimination only.

\hypertarget{bs}{%
\subsubsection{Brier Score (BS)}\label{bs}}
The BS \citep{graf1999} is defined as \begin{align}\label{eqn:bs}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
\end{align} where \(\delta_{i}=1\) shows individuals who have experienced the event, \(N\) represents the number of samples
in our dataset over which we calculate \(BS(t)\), \(T_{i}\) is the survival or censoring time of an individual, \(\widehat{G}(t)=P[c>t]\) is a non-parametric estimate of the censoring distribution and \(c\) is censoring time. The BS provides a score that accounts for the
information loss because of censoring. Once we fix our \(t\) of interest, the individuals in the dataset can be divided into three
groups. Individuals who experienced the event before \(t\) are present in the first term of the equation. The second term of the
equation includes individuals who experience the event or are censored after \(t\). Those censored before \(t\) (the remaining
individuals) are accounted for by the IPCW adjustment (\(G(\cdot)\)) present on both terms.

@@

The Integrated Brier Score (IBS) is a function of the BS (\ref{eqn:bs}) \citep{graf1999}, which is defined as \begin{align*}
IBS(t)=\int_{t}^{t_{max}}BS(t)dw(t),
\end{align*} where $w(t)=\frac{t}{t_{max}}$ and $t_{max}$ is the upper bound for the survival times of interest. As the IBS is calculated for a range of follow-up-times,
it is a useful metric to track overall performance across all of follow-up time during cross-validation. We use the
IPA score to get an estimate of performance at each follow-up time for our studies.

@@

The IPA is a function of time, based on the BS. The IPA score is given by \begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where \(BS_{model}(t)\) represents the BS over time \(t\) for the model of interest and \(BS_{null}(t)\) represents
the BS if we use an unadjusted Kaplan-Meier (KM) curve as the prediction for all observations \citep{kattan2018index}. The
IPA score has an upper bound of one, where positive values show an increase in performance over the null model and
negative values show that the null model performs better. As an extension of BS, the IPA score assesses both model calibration
and discrimination and show how performance changes over follow-up time \citep{graf1999} \citep{kattan2018index}.

@@

R \citep{Rsoft} and Python \citep{py} are used to evaluate methods from both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg} and the CBLR model using the \textbf{casebase} package \citep{cbpkg}. Both DeepSurv and DeepHit are fit using \textbf{pyCox}
\citep{lee2018DeepHit}. We made the components of CBNN using the \textbf{casebase} package \citep{cbpkg} for the sampling step and the
\textbf{keras} package \citep{keras} for our neural network architecture. We use the \textbf{simsurv} package \citep{simsurv} for our simulation studies
and \textbf{flexsurv} \citep{flexsurv} to fit a flexible baseline hazard using splines for our complex simulation. The \textbf{riskRegression} package
\citep{riskRegression} is used to get the IPA and $AUC_{IPCW}$. We modify the \textbf{riskRegression} package to be used with any user supplied
risk function \(F\). We use the \textbf{reticulate} package \citep{reticulate} to run both R based methods and Python based methods on the same seed.

@@

\caption{Performance of each model in the complex simulation (A, E), multiple myeloma (MM) case study (B, F), free light chain (FLC) case study (C, G)
and prostate cancer (Prostate) case study (D, H). The first row shows the IPA for each model in each study over follow-up time. Negative values mean
the model performs worse than the null model and positive values mean the model performs better. The second row shows the $AUC_{IPCW}$ for
each model in each study over follow-up time, where higher is better. Each model-specific metric in each study shows a 95\% confidence interval over 100 iterations. Metrics
are shown for six models: Case-Base with Logistic Regression (CBLR), Case-Base Neural Network (CBNN), Cox Proportional Hazard (Cox), DeepHit
and DeepSurv. The Kaplan-Meier (KM) model serves as a baseline, predicting the average curve for all individuals. CBLR and Cox have near identical
performance, resulting in curves that overlap. The Optimal model (a CBLR model with the exact interaction terms and baseline hazard specified) shows
the best performance we can expect on the simulated data.}\label{fig:megaPlot}
\end{figure}

@@

We simulate data to evaluate the performance of CBNN in comparison with existing regression (Cox, CBLR) and neural network (DeepHit, DeepSurv)
methods. We specify a linear combination of each covariate as the linear predictor in the regression-based methods (Cox, CBLR), which contrasts
with neural network approaches that allow for approximations of non-linear interactions. The simulated data are based on a complex baseline hazard with time-varying
interactions, each with 10\% random censoring. We simulate three covariates for 5000 individuals:

\[
z_{1} \sim \textrm{Bernoulli}(0.5) \qquad \qquad %technically a control and treatment group
z_{2} \sim \begin{cases}
 N(0,0.5) & \textrm{if } z_{1}=0\\ 
 N(1,0.5) & \textrm{if } z_{1}=1
\end{cases} \qquad \qquad
z_{3} \sim N(1,0.5).
\]
In addition to the methods described above, we include the exact functional form of the covariates in a CBLR model (referred to as Optimal for simplicity)
in the complex simulation. We obtain confidence intervals by conducting 100 bootstrap re-samples on the training data. We keep 15\% of the data for testing before
hyperparameter selection. We use 15\% of the remaining data for validation and the rest is reserved for training. We predict risk functions for individuals in
the test set, which are used to calculate our IPA and $AUC_{IPCW}$.

@@

Figure \ref{fig:megaPlot} A, E and Table \ref{tab:megaTable} shows the performance over time on a test set. The Optimal model acts as a reference for ideal performance on the simulated data. For discrimination, the Optimal model performs best,
followed by CBNN, DeepHit, DeepSurv and the linear models (Figure \ref{fig:megaPlot} E). For both   To obtain a more realistic performance assessment, we compared
models in three case studies with a time-to-event outcome.

\begin{table}[!htbp]
\caption{Performance at certain percentages of follow-up time in the complex simulation (A), multiple myeloma (MM) case study (B), free light chain (FLC)
case study (C) and prostate cancer (Prostate) case study (D). Each table shows performance for each method in each study at $25\%$, $50\%$, $75\%$
and $100\%$ of follow-up time. These tables are included to
provide exact measures of performance. The models of interest are case-base with logistic regression (CBLR), Cox, Case-Base Neural Network (CBNN),
DeepHit, DeepSurv, and Optimal (in the complex simulation). The best score at each percent of follow-up time is highlighted in bold. If the average performance
is tied, then all tied values are highlighted.}
\label{tab:megaTable}
